---
title: "Player Analysis: Sparse PCA"
author: "Tirdod Behbehani"
date: "19.12.2024"
format: 
  pdf:
    keep-tex: true
    include-in-header: 
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
         \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
---

# 1. Introduction
This document explores player data using K-means clustering and PCA to understand if there is any structure in the data that can inform market value predictions. 

---

# Data Import and Preparation

## Load Required Libraries
```{r setup, include=TRUE}
# Load libraries
library(tidyverse)    # Data manipulation and visualization
library(cluster)      # Clustering algorithms
library(factoextra)   # Visualization for clustering
library(broom)        # Tidy model outputs
library(knitr)        # Pretty tables
library(dplyr)
library(gridExtra) # multiple plotting with ggplot2
library(mclust) # clustering with mixture model
library(gclus) # Clustering Graphics package
library(car)
library(corrplot)      # Install corrplot if not already installed
library(ggcorrplot)    # Install ggcorrplot if not already installed
library(elasticnet)
library(PMA)
library(cowplot)
library(caret)
```

## Import data
```{r}

#Import data
per90s_with_names <- read_csv('merged_90s_cleaned.csv')#[ , -1]
totals_with_names <- read_csv('merged_total_cleaned.csv')#[ , -1]
 
per90s <- per90s_with_names[, -(1:8)]
totals <- totals_with_names[, -(1:8)]


```

look at correlation matrix
```{r}
# Compute the correlation matrix (for totals)
cor_matrix <- cor(per90s, use = "complete.obs")  

# Find highly correlated pairs (threshold = 0.8)
threshold <- 0.75
high_cor_pairs <- which(abs(cor_matrix) > threshold, arr.ind = TRUE)

# Exclude self-correlations (correlation of a variable with itself)
high_cor_pairs <- high_cor_pairs[high_cor_pairs[, 1] != high_cor_pairs[, 2], ]

# Create a df of highly correlated pairs
high_cor_df <- data.frame(
  Var1 = rownames(cor_matrix)[high_cor_pairs[, 1]],
  Var2 = colnames(cor_matrix)[high_cor_pairs[, 2]],
  Correlation = cor_matrix[high_cor_pairs]
)

# Remove duplicate pairs 
high_cor_df <- high_cor_df[!duplicated(t(apply(high_cor_df, 1, sort))), ]

high_cor_df <- high_cor_df[order(-high_cor_df$Correlation), ]

# Print the highly correlated pairs
print("Highly Correlated Pairs:")
print(high_cor_df)

#now we remove highly correlated columns from the 90s df:
per90s_clean <- per90s[, !(names(per90s) %in% c("short_pass_att_per90", "medium_pass_att_per90", "long_pass_att_per90","all_pass_att_per90",
                              "pk_gls_per90", "pk_att_per90","npxg_minus_xg_per90","npxg_per90","take_ons_att_per90",
                              "sca_per90","sca_live_ball_pass_per90", "sca_dead_ball_pass_per90", "sca_shot_per90",
                              "sca_fouled_per90", "sca_defensive_per90",
                              "sca_take_on_per90","gca_take_on_per90",
                              "gca_per90","gca_live_ball_pass_per90", "gca_dead_ball_pass_per90", "gca_shot_per90",
                              "gca_fouled_per90", "gca_defensive_per90",
                              "tkl_att_per90","total_pass_cmp%","aerial_duels_att_per90","touches_def_pen_area_per90",
                              "touches_pen_area_per90",
                              "shots_on_target_per90","all_pass_cmp_per90","xa_per90","goals_per_shot_on_target",
                              "drbl_tkl_attempted_per90",
                              "npxg_per_shot_per90","blocks_per90","goals_per_shot","average_shot_distance_meters",
                              "crosses_per90"
                              ))]

# Compute the correlation matrix (for totals)
cor_matrix <- cor(per90s_clean, use = "complete.obs")  

# Find highly correlated pairs (threshold = 0.8)
threshold <- 0.75
high_cor_pairs <- which(abs(cor_matrix) > threshold, arr.ind = TRUE)

# Exclude self-correlations (correlation of a variable with itself)
high_cor_pairs <- high_cor_pairs[high_cor_pairs[, 1] != high_cor_pairs[, 2], ]

# Create a df of highly correlated pairs
high_cor_df <- data.frame(
  Var1 = rownames(cor_matrix)[high_cor_pairs[, 1]],
  Var2 = colnames(cor_matrix)[high_cor_pairs[, 2]],
  Correlation = cor_matrix[high_cor_pairs]
)

# Remove duplicate pairs 
high_cor_df <- high_cor_df[!duplicated(t(apply(high_cor_df, 1, sort))), ]

high_cor_df <- high_cor_df[order(-high_cor_df$Correlation), ]

# Print the highly correlated pairs
print("Highly Correlated Pairs:")
print(high_cor_df)

```

## perform train and test split and preprocess data


```{r split-and-preprocess}

# set.seed(42)
# 
# # 1) Split on raw per90s_clean
# per90s_index   <- createDataPartition(1:nrow(per90s_clean), p = 0.8, list = FALSE)
# train_per90 <- per90s_clean[ train_idx, ]
# test_per90  <- per90s_clean[-train_idx, ]

```



check if there are any variables with suspicious distributions
``` {r}
library(e1071)  # for skewness and kurtosis

# Flag variables with problematic distributions
check_distributions <- function(df) {
  results <- data.frame(
    metric = colnames(df),
    skewness = apply(df, 2, skewness, na.rm = TRUE),
    kurtosis = apply(df, 2, kurtosis, na.rm = TRUE),
    sd = apply(df, 2, sd, na.rm = TRUE)
  )

  # Set thresholds for flagging
  results$flag_skewness <- abs(results$skewness) > 2
  results$flag_kurtosis <- results$kurtosis > 7
  results$flag_low_variance <- results$sd < 0.01
  results$flag_high_sd <- results$sd > 3

  # Combine flags
  results$red_flag <- with(results, flag_skewness | flag_kurtosis | flag_low_variance | flag_high_sd)

  return(results[results$red_flag == TRUE, ])
}

# Run on your scaled training set
red_flags <- check_distributions(per90s_clean)

# Print flagged metrics
print(red_flags)
```

``` {r}

# turn the errors leading the shots to just flag the 10% of players with the most errors
# Step 1: Compute 90th percentile threshold
threshold_90 <- quantile(per90s_clean$error_leading_to_shot_per90, probs = 0.90, na.rm = TRUE)

# Step 2: Create flag column for top 10%
per90s_clean$high_error_flag <- per90s_clean$error_leading_to_shot_per90 >= threshold_90
per90s_clean$error_leading_to_shot_per90 <- NULL

# test_per90$high_error_flag  <- test_per90$error_leading_to_shot_per90  >= threshold_90
# test_per90$error_leading_to_shot_per90  <- NULL

## 3) Compute 99th‐pct caps on train & apply to both

# the three columns you intend to winsorize:
want_to_cap <- c(
  "crosses_into_penalty_area_per90",
  "carries_into_pen_area_per90",
  "aerial_duels_won_per90"
)

# Only keep the ones that actually exist in train_per90
cap_cols <- intersect(want_to_cap, names(per90s_clean))

for(col in cap_cols) {
  # 1) compute the 99th‐pctile 
  cap <- quantile(per90s_clean[[col]], 0.99, na.rm=TRUE)
  
  # 2) in train & test, any value above cap → set to cap
  per90s_clean[[col]][ per90s_clean[[col]] > cap ] <- cap
  #test_per90[[col]][  test_per90[[col]] > cap ] <- cap
}

# 4) Fit scaler on train only, then transform both
preProc_90s <- preProcess(per90s_clean, method = c("center","scale"))
all_scaled  <- predict(preProc_90s, per90s_clean)
#test_scaled   <- predict(preProc_90s, test_per90)

# 5) Quick check
summary(all_scaled)
#summary(test_scaled)

```


## Tune and fit the sparse pca model
```{r}

# Use the training dataset directly
X <- as.matrix(all_scaled)  # Training data (no further splitting)

# Hyperparameter grids
K_values <- c(1:7)  # Number of components to test
lambda_values <- seq(0.05, 0.2, by = 0.05)  # Sparsity penalty values

# A function to compute explained variance ratio
explained_variance_ratio <- function(X_original_centered, loadings, scores) {
  X_reconstructed <- scores %*% t(loadings)
  
  # Compute the variance explained by reconstructed data
  var_explained <- sum(apply(X_reconstructed, 2, var))
  
  # Compute the total variance in the original centered data
  var_total <- sum(apply(X_original_centered, 2, var))
  
  # Return the ratio of explained variance
  var_explained / var_total
}

# Center the training dataset
all_mean <- colMeans(X)
X_centered <- sweep(X, 2, all_mean, FUN = "-")

# Compute the covariance matrix of the centered training data
#cov_matrix <- cov(X_centered)

# Loop over hyperparameter combinations
results <- data.frame(K = integer(), lambda = numeric(), ev_ratio = numeric())

for (K in K_values) {
  for (lambda in lambda_values) {
    
    # Fit Sparse PCA at (K, λ)
    fit <- spca(X_centered,
        K        = K,
        type     = "predictor",
        sparse   = "penalty",
        para     = rep(lambda, K))
    loadings <- fit$loadings
    
    scores <- X_centered %*% loadings  # Scores for the training set

    # Compute explained variance ratio
    ev <- explained_variance_ratio(X_centered, loadings, scores)
    
    # Store the results
    results <- rbind(results, data.frame(K = K, lambda = lambda, ev_ratio = ev))
  }
}

# Find the best (K, lambda) based on the highest explained variance
best_params <- results[which.max(results$ev_ratio), ]
print(best_params)

# **Now** re-fit your final model once, outside the loops
best_fit <- spca(X_centered,
                  K      = best_params$K,
                  type   = "predictor",
                  sparse = "penalty",
                  para   = rep(best_params$lambda, best_params$K))

```


taking the optimal lambda from above, now let's look at the explained variance for up to 25 principal components
```{r}

# lambda <- best_params$lambda
# 
# max_vars <- 25
# 
# ev_ratios <- numeric(max_vars)
# 
# for (K in 1:max_vars) {
# 
#   fit <- spca(X_centered, K = K, type = "predictor",
#               sparse = "penalty", para = rep(lambda, K))
#   ## Pull loadings from this fit, not the best_fit
#   loadings <- fit$loadings
#   
#   scores <- X_centered %*% loadings
#   
#   # Compute explained variance ratio
#   X_reconstructed <- scores %*% t(loadings)
#   var_explained <- sum(apply(X_reconstructed, 2, var))
#   var_total <- sum(apply(X_centered, 2, var))
#   
#   ev_ratio <- var_explained / var_total
#   ev_ratios[K] <- ev_ratio
#   
#   # If we reach or exceed 100% (or very close), break out
#   if (ev_ratio >= 1) {
#     break
#   }
# }
# 
# # Filter out any trailing zeros if we broke early
# ev_ratios <- ev_ratios[ev_ratios > 0]
# 
# # Create a data frame for plotting
# df <- data.frame(
#   K = 1:length(ev_ratios),
#   ev_pct = ev_ratios * 100
# )
# 
# ggplot(df, aes(x = K, y = ev_pct)) +
#   geom_line(color = "steelblue", size = 1) +
#   geom_point(size = 3, color = "steelblue") +
#   # Add dashed lines at 70%, 80%, 90% for reference
#   geom_hline(yintercept = c(70,80,90), linetype = "dashed", color = "red") +
#   annotate("text", x = max(df$K) + 0.3, y = 70, label = "70%", vjust = -0.5, color = "red") +
#   annotate("text", x = max(df$K) + 0.3, y = 80, label = "80%", vjust = -0.5, color = "red") +
#   annotate("text", x = max(df$K) + 0.3, y = 90, label = "90%", vjust = -0.5, color = "red") +
#   scale_x_continuous(breaks = seq(1, max(df$K), by = 1)) +
#   scale_y_continuous(limits = c(0,100)) +
#   labs(
#     x = "Number of Components (K)",
#     y = "Cumulative Explained Variance (%)",
#     title = sprintf("Cumulative Explained Variance by SPCA (lambda = %.2f)", lambda)
#   ) +
#   theme_minimal() +
#   theme(
#     panel.grid.minor = element_blank(),
#     plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
#     axis.title = element_text(size = 11, face = "bold"),
#     axis.text = element_text(size = 10)
#   )
# 
# # Round to two decimals
# df_table <- df %>% 
#   dplyr::mutate(ev_pct = round(ev_pct, 2)) %>% 
#   dplyr::rename(
#     NumComponents      = K,
#     CumulativeVariance = ev_pct
#   )
# 
# # 2) Or a prettier table in RMarkdown
# knitr::kable(
#   df_table,
#   caption = "Number of Components vs. Cumulative Explained Variance (%)",
#   col.names = c("K", "Cum. Var. (%)"),
#   digits = 2
# )

```


```{r varimax-rotation}
# 1) Grab the raw sparse‐PCA loadings for PCs 1–7
rawL <- best_fit$loadings

# 2) Apply an orthogonal Varimax rotation
rot     <- varimax(rawL)
rotL    <- rot$loadings        # rotated loadings matrix (p × 5)


# 2) Build a data.frame with pairs of columns
compare <- data.frame(
  Metric    = rownames(rawL),
  Raw_PC1   = round(rawL[,1], 3),
  Rot_PC1   = round(rotL[,1], 3),
  Raw_PC2   = round(rawL[,2], 3),
  Rot_PC2   = round(rotL[,2], 3),
  Raw_PC3   = round(rawL[,3], 3),
  Rot_PC3   = round(rotL[,3], 3),
  Raw_PC4   = round(rawL[,4], 3),
  Rot_PC4   = round(rotL[,4], 3),
  Raw_PC5   = round(rawL[,5], 3),
  Rot_PC5   = round(rotL[,5], 3),
  Raw_PC6   = round(rawL[,6], 3),
  Rot_PC6   = round(rotL[,6], 3),
  Raw_PC7   = round(rawL[,7], 3),
  Rot_PC7   = round(rotL[,7], 3),
  check.names = FALSE
)

# 3) Print with knitr::kable
library(knitr)
kable(compare,
      caption = "Unrotated vs Varimax‐Rotated SPCA Loadings (PC1–PC7)",
      align = c("l", rep("r", 14)),
      digits = 3)

```



## analyze principal component features 

rotated
``` {r}

# 1) Extract the raw loadings for PCs 1:K_rot
rawL <- best_fit$loadings[, 1:K]

# 2) Apply orthogonal Varimax rotation
rot      <- varimax(rawL)
rotL     <- rot$loadings        # rotated loadings matrix (p × K_rot)


# 3) compute the rotated scores: n×p  %*%  p×7  =  n×7
scores_rot <- X_centered %*% rotL

# Flip PC1 so that higher = more defender-type
rotL[,1]       <- -rotL[,1]
scores_rot[,1] <- -scores_rot[,1]

# Flip PC5 so that higher = more dribbler-type
rotL[,5]       <- -rotL[,5]
scores_rot[,5] <- -scores_rot[,5]

# Flip PC6 so that higher = more goalscorer-type
rotL[,6]       <- -rotL[,6]
scores_rot[,6] <- -scores_rot[,6]

# 4) Print top 10 variables for each rotated PC
L <- rotL
K <- ncol(L)

for (j in 1:K) {
  cat("Top 10 variables for Rotated PC", j, ":\n")
  
  # order by absolute loading, descending
  abs_loads <- sort(abs(L[, j]), decreasing = TRUE)
  
  # pick the top 10
  top_vars <- names(abs_loads)[1:10]
  
  # build a small data.frame
  df_top <- data.frame(
    Variable = top_vars,
    Loading  = round(L[top_vars, j], 3),
    row.names = NULL
  )
  
  print(df_top)
  cat("\n")
}

```


``` {r}
# 4) name your archetypes in the same order as the PCs
archetypes <- c(
  "Anchor",
  "Ball_Progressor",
  "Ball_Winner",
  "Wide_Playmaker",
  "Dribbler",
  "Finisher",
  "Advanced_Playmaker"
)

player_scores <- as_tibble(scores_rot) %>%
  setNames(archetypes) %>%
  mutate(
    player   = per90s_with_names$player,
    team     = per90s_with_names$team,
    season   = per90s_with_names$season,
    league   = per90s_with_names$league,
    nation   = per90s_with_names$nation,
    pos      = per90s_with_names$pos,
    age      = per90s_with_names$age,
    `90s`    = per90s_with_names$`90s`    # <— backticks on both sides
  ) %>%
  select(player, team, season, league, nation, pos, age, `90s`, everything())

# 5) take a look
player_scores %>% slice(1:10)

player_scores_norm <- player_scores %>%
  # for each archetype column, do: (x – min) / (max – min) * 100
  mutate(across(all_of(archetypes),
                ~ (. - min(., na.rm = TRUE)) /
                  (max(., na.rm = TRUE) - min(., na.rm = TRUE)) * 100))

# Inspect the normalized scores
player_scores_norm %>% slice(1:10)

```


draw radar / spider plots
# ``` {r}
# 
# library(dplyr)
# 
# # Pick a single player (or filter multiple)
# to_plot <- player_scores_norm %>% 
#   filter(player == "Virgil van Dijk",
#          season == "2019-2020")
# 
# # Extract just the archetype‐score columns
# axes <- c("Anchor",
#           "Finisher",
#           "Playmaker",
#           "Dribbler",
#           "Wide_Playmaker",
#           "Ball_Winner",
#           "Ball_Progressor")
# df_raw <- to_plot[ , axes ]
# 
# # Build an fmsb‐friendly DF: first row = max across axes, second = min
# max_row <- rep(100, length(axes))
# min_row <- rep(0,   length(axes))
# radar_df <- rbind(max_row, min_row, df_raw)
# colnames(radar_df) <- axes
# 
# ```


``` {r}

plot_player_archetype <- function(player_name, season_name,
                                  axes = c("Anchor",
                                           "Finisher",
                                           "Advanced_Playmaker",
                                           "Dribbler",
                                           "Wide_Playmaker",
                                           "Ball_Winner",
                                           "Ball_Progressor"),
                                  col_border = "#CE1141",
                                  col_fill   = alpha("#CE1141", .3)) {
  # 1) look up the row in your already‐built scores & meta tables
  to_plot <- player_scores_norm %>%
    filter(player == player_name, season == season_name)
  if (nrow(to_plot) != 1) {
    stop("Expected exactly one match for that player & season.")
  }
  md <- per90s_with_names %>%
    filter(player == player_name, season == season_name) %>%
    slice(1)
  
  # 2) build fmsb radar_df
  df_raw  <- to_plot[ , axes]
  radar_df <- rbind(
    rep(100, length(axes)),
    rep(0,   length(axes)),
    df_raw
  )
  colnames(radar_df) <- axes
  
  # 3) set up plotting region
  oldpar <- par(
    mar       = c(0, 0, 0, 0),
    oma       = c(0, 0, 0, 0),
    plt       = c(0.08, 0.9, 0, 0.88),
    font.axis = 2
  )
  
  # 4) blank spider
  radarchart(
    radar_df,
    axistype   = 0,
    seg        = 4,
    pcol       = col_border,
    pfcol      = col_fill,
    plwd       = 2,
    plty       = 1,
    cglcol     = "grey10",
    cglty      = 1,
    cglwd      = 0.8,
    axislabcol = "grey10",
    vlabels    = rep("", length(axes)),
    vlcex      = 1.1,
    calcex     = 0.7
  )
  
  # 5) bolded archetype labels
  N      <- length(axes)
  angles <- seq(0, 2*pi, length.out = N+1)[- (N+1)] + pi/2
  r0     <- 1.10
  lbls   <- gsub("_", " ", axes)
  lbls[lbls=="Advanced Playmaker"] <- "Advanced\nPlaymaker"
    lbls[lbls=="Wide Playmaker"] <- "Wide\nPlaymaker"
  for (i in seq_along(lbls)) {
    xi <- cos(angles[i]) * r0
    yi <- sin(angles[i]) * r0
    # your custom nudges
    if (lbls[i]=="Anchor")         yi <- yi + 0.05
    if (lbls[i]=="Finisher")       { xi<-xi-0.1; yi<-yi+0.1 }
    if (lbls[i]=="Ball Progressor"){ xi<-xi+0.2; yi<-yi+0.1 }
    if (lbls[i]=="Advanced\nPlaymaker") {
      xi <- xi - 0.15
      yi <- yi - 0.15
    }
    if (lbls[i]== "Ball Winner")    { xi<-xi+0.18; yi<-yi-0.15 }
    if (lbls[i] == "Dribbler") yi<-yi-0.05
      if (lbls[i] == "Wide\nPlaymaker") yi<-yi-0.11
    text(xi, yi, lbls[i], font=2, cex=1.1, adj=c(0.5,0.5))
  }
  
  # 6) 25–100% ticks on every spoke, then one 0% in the middle
  pct_vals <- c(0.25, 0.50, 0.75, 1.00)  # drop the 0
  for (i in seq_along(angles)) {
    for (p in pct_vals) {
      # bump 25% and 50% rings a little so they don’t overlap:
      if (abs(p - 0.25) < 1e-8) {
        p_adj <- p + 0.075
      } else if (abs(p - 0.50) < 1e-8) {
        p_adj <- p + 0.03
      } else if (abs(p - 0.75) < 1e-8) {
        p_adj <- p - 0.02
      } else {
        p_adj <- p - 0.07  # push the outer ring labels slightly in
      }
      x <- cos(angles[i]) * p_adj
      y <- sin(angles[i]) * p_adj
      text(x, y, paste0(p*100, "%"),
           cex   = 0.7,
           col   = "grey10",
           font  = 2,
           adj   = c(0.5, 0.5))
    }
  }

  # now add a single “0%” at the center
  text(0, 0, "0%", cex = 0.7, col = "grey10", font = 2, adj = c(0.5, 0.5))
  
  # 7) three‐line bold title
  main_expr <- bquote(
    atop(
      bold(.(player_name)),
      atop(
        .(paste0(md$team, " (", md$season, ")")),
        "Age:"~.(md$age)
      )
    )
  )
  title(main = main_expr, adj=0, line=0,
        cex.main=1.6, col.main="#333333", font.main=1)
  
  par(oldpar)
}

```

``` {r}

# ── Usage ─────────────────────────────────────────────────────────────────────
plot_player_archetype("Virgil van Dijk", "2022-2023")

plot_player_archetype("Cristiano Ronaldo", "2019-2020")

plot_player_archetype("Rodri", "2021-2022")

plot_player_archetype("Lionel Messi", "2019-2020")

plot_player_archetype("Lionel Messi", "2020-2021")

plot_player_archetype("Lionel Messi", "2021-2022")

plot_player_archetype("Lionel Messi", "2022-2023")



```
